Prominent Factors that Influenced Canadian Egg Prices in 2024

### 2024-12-04 paper/paper.qmd
paper

The substantial price discrepancies observed across different vendors point to the possible presence of **oligopolistic behavior** within the egg market. In an oligopoly, a small number of firms exert significant market power, enabling them to influence prices independently of other factors. This could explain the significant fluctuations in egg prices observed between vendors, even when other variables, such as egg size or farming method, remain constant. The concentration of market power among a few dominant vendors could limit competition, allowing them to adjust prices in ways that benefit them, potentially at the expense of consumers.

Overall, this study underscores the importance of considering **vendor influence** when analyzing egg prices. It challenges the conventional view that factors such as farming practices and egg characteristics are the primary drivers of price. The findings suggest that a deeper understanding of vendor behavior and market concentration is essential for policymakers and consumers alike. Future research should explore these dynamics further, particularly by examining regional market structures, vendor-specific pricing strategies, and potential regulatory interventions aimed at improving market transparency and competition.

@@

First, the assumption of **linearity** between the independent variables (e.g., vendor, egg size, farming method) and the dependent variable (egg price) may not fully capture the true relationships. If these relationships are non-linear, the model could provide biased estimates. Additionally, multicolinearity—where variables such as vendor type and farming method are highly correlated—could distort the coefficient estimates, making it difficult to isolate the independent effects of each factor on price.

@@

Lastly, while the analysis identifies vendor as a significant determinant of egg prices, the model may not have fully captured the complexity of **vendor-specific pricing strategies** or **regional market dynamics**, which could vary beyond the variables included in the regression.

@@

Another promising direction for future research is to explore the impact of marketconcentration. This study suggests that the Canadian egg market may exhibit oligopolistic characteristics, but it does not fully explore how market concentration influences pricing in a dynamic way. Analyzing the relationship between market share, competition, and pricing could offer deeper insights into whether fewer vendors in the market lead to higher or more stable prices, and how this might affect consumers. Comparative studies across other food sectors or countries with different market structures could provide valuable lessons.

@@

Finally, this study focused on a specific set of variables and used a **multivariate linear** regression model. Future work could apply alternative modeling techniques, such as machine learning or non-linear models, to capture more complex relationships between pricing and the independent variables. Expanding the dataset to include a larger sample size, longer time periods, or more granular geographic data could also strengthen the robustness and generalizability of the results.

@@

While addressing the structural issues of Canada's grocery market—such as its oligopolistic tendencies—remains a challenging task, there are practical steps consumers can take to reduce their grocery bills in the short term. One effective strategy is to purchase store brands, like No Name, which tend to offer more affordable alternatives without compromising quality. In addition, consumers can incorporate more plant-basedmeals into their diets, as plant-based foods like beans, lentils, and grains are often more affordable than animal products. Another cost-saving approach is to buy in bulk, particularly for non-perishable items such as rice, beans, and pasta, which can reduce costs over time. **Meal planning** is also a valuable strategy, helping consumers buy only what they need, reduce food waste, and avoid impulse purchases—ultimately contributing to lower grocery costs \[@citeHughes\].

@@

The data collection for this study was based on screen scraping of the website user interfaces (UIs) of eight well-known vendors in Canada, as outlined by Project Hammer \[\@citeDate\]. The screen scraping technique involved programmatically extracting price information for eggs and related product variables from the vendors' online storefronts. This data was specifically gathered through the "in-store pickup" option, as this method typically provides accurate pricing and product availability, reflecting actual prices that consumers would pay when purchasing eggs for pickup rather than delivery. This method is often chosen because it minimizes discrepancies that can arise due to delivery fees or location-specific pricing.While this approach provided valuable data, there are limitations to screen scraping, particularly in the context of price accuracy and completeness. Websites often display product information in a way that is optimized for visual presentation rather than data extraction, which can result in incomplete or inconsistent data. Furthermore, vendor websites are frequently updated, and changes to UI elements or the underlying structure of web pages could lead to the loss of important data or inaccurate price representations.A more robust and comprehensive method of data collection would be the use of internal APIs provided by grocery retailers, which are often more structured and reliable than publicly accessible websites. APIs allow direct access to the retailer's product catalog, offering a more complete and accurate set of product options, prices, and related data, often including various filters and variables not visible on the website UI. In contrast to screen scraping, APIs are typically designed to deliver raw, structured data that can be easily queried, making it a more consistent and efficient data source. For instance, APIs would list not only the eggs available for in-store pickup, but also the full range of products that could be purchased, including promotions, bulk discounts, and regional variations that might not be represented on the website’s UI.Using internal APIs would also mitigate the risk of data loss due to website redesigns or technical issues. A study by Choudhury et al. (2017) on API-based data collection in e-commerce found that using APIs led to more accurate price tracking and more complete product information compared to web scraping, as APIs provide direct access to a vendor’s backend systems [@citeChoudhury]. Moreover, internal APIs often provide real-time data updates, which is particularly important when tracking fluctuating commodity prices like eggs.However, while using APIs presents numerous advantages, it is important to note that this method requires explicit access permissions from the vendors themselves. Retailers may impose limitations on API access, or charge for this data, which could increase the complexity and cost of the data collection process. Additionally, not all grocers may offer open or public APIs, limiting the scope of this method.In summary, while screen scraping of website UIs provided a useful data source for this study, using internal APIs would offer a more robust and scalable approach to data collection, providing more accurate, complete, and timely data for price analysis. Further research in this area could explore how API access could be standardized across the grocery industry to support more comprehensive price tracking and market analysis.

>>>>

The substantial price discrepancies observed across different vendors point to the possible presence of monopolistic behavior within the egg market. In an oligopoly, a small number of firms exert significant market power, enabling them to influence prices independently of other factors. This could explain the significant fluctuations in egg prices observed between vendors, even when other variables, such as egg size or farming method, remain constant. The concentration of market power among a few dominant vendors could limit competition, allowing them to adjust prices in ways that benefit them, potentially at the expense of consumers.

Overall, this study underscores the importance of considering vendor influence when analyzing egg prices. It challenges the conventional view that factors such as farming practices and egg characteristics are the primary drivers of price. The findings suggest that a deeper understanding of vendor behavior and market concentration is essential for policymakers and consumers alike. Future research should explore these dynamics further, particularly by examining regional market structures, vendor-specific pricing strategies, and potential regulatory interventions aimed at improving market transparency and competition.

@@

First, the assumption of linearity between the independent variables (e.g., vendor, egg size, farming method) and the dependent variable (egg price) may not fully capture the true relationships. If these relationships are non-linear, the model could provide biased estimates. Additionally, multicolinearity—where variables such as vendor type and farming method are highly correlated—could distort the coefficient estimates, making it difficult to isolate the independent effects of each factor on price.

@@

Lastly, while the analysis identifies vendor as a significant determinant of egg prices, the model may not have fully captured the complexity of vendor-specific pricing strategies or regional market dynamics, which could vary beyond the variables included in the regression.

@@

Another promising direction for future research is to explore the impact of market concentration. This study suggests that the Canadian egg market may exhibit oligopolistic characteristics, but it does not fully explore how market concentration influences pricing in a dynamic way. Analyzing the relationship between market share, competition, and pricing could offer deeper insights into whether fewer vendors in the market lead to higher or more stable prices, and how this might affect consumers. Comparative studies across other food sectors or countries with different market structures could provide valuable lessons.

@@

Finally, this study focused on a specific set of variables and used a multivariate linear regression model. Future work could apply alternative modeling techniques, such as machine learning or non-linear models, to capture more complex relationships between pricing and the independent variables. Expanding the dataset to include a larger sample size, longer time periods, or more granular geographic data could also strengthen the robustness and generalization of the results.

@@

While addressing the structural issues of Canada's grocery market—such as its monopolistic tendencies—remains a challenging task, there are practical steps consumers can take to reduce their grocery bills in the short term. One effective strategy is to purchase store brands, like No Name, which tend to offer more affordable alternatives without compromising quality. In addition, consumers can incorporate more plant-based meals into their diets, as plant-based foods like beans, lentils, and grains are often more affordable than animal products. Another cost-saving approach is to buy in bulk, particularly for non-perishable items such as rice, beans, and pasta, which can reduce costs over time. Meal planning is also a valuable strategy, helping consumers buy only what they need, reduce food waste, and avoid impulse purchases—ultimately contributing to lower grocery costs [@citeHughes].

@@

The data collection for this study was based on screen scraping of the website user interfaces (UIs) of eight well-known vendors in Canada, as outlined by Project Hammer [@citeData]. The screen scraping technique involved programmatically extracting price information for eggs and related product variables from the vendors' online storefronts. This data was specifically gathered through the "in-store pickup" option, as this method typically provides accurate pricing and product availability, reflecting actual prices that consumers would pay when purchasing eggs for pickup rather than delivery. This method is often chosen because it minimizes discrepancies that can arise due to delivery fees or location-specific pricing.While this approach provided valuable data, there are limitations to screen scraping, particularly in the context of price accuracy and completeness. Websites often display product information in a way that is optimized for visual presentation rather than data extraction, which can result in incomplete or inconsistent data. Furthermore, vendor websites are frequently updated, and changes to UI elements or the underlying structure of web pages could lead to the loss of important data or inaccurate price representations.A more robust and comprehensive method of data collection would be the use of internal APIs provided by grocery retailers, which are often more structured and reliable than publicly accessible websites. APIs allow direct access to the retailer's product catalog, offering a more complete and accurate set of product options, prices, and related data, often including various filters and variables not visible on the website UI. In contrast to screen scraping, APIs are typically designed to deliver raw, structured data that can be easily queried, making it a more consistent and efficient data source. For instance, APIs would list not only the eggs available for in-store pickup, but also the full range of products that could be purchased, including promotions, bulk discounts, and regional variations that might not be represented on the website’s UI.Using internal APIs would also mitigate the risk of data loss due to website redesigns or technical issues. A study by Choudhury et al. (2017) on API-based data collection in e-commerce found that using APIs led to more accurate price tracking and more complete product information compared to web scraping, as APIs provide direct access to a vendor’s backend systems [@citeChoudhury]. Moreover, internal APIs often provide real-time data updates, which is particularly important when tracking fluctuating commodity prices like eggs.However, while using APIs presents numerous advantages, it is important to note that this method requires explicit access permissions from the vendors themselves. Retailers may impose limitations on API access, or charge for this data, which could increase the complexity and cost of the data collection process. Additionally, not all grocers may offer open or public APIs, limiting the scope of this method.In summary, while screen scraping of website UIs provided a useful data source for this study, using internal APIs would offer a more robust and scalable approach to data collection, providing more accurate, complete, and timely data for price analysis. Further research in this area could explore how API access could be standardized across the grocery industry to support more comprehensive price tracking and market analysis.
